The Ranked Nearest Neighbour Interchange space

### 2019-06-13 paper.tex
bound for length of path when deleting one taxon from all trees

>>>>

%This lemma might be central  for induction proofs in RNNI space
%Is it correct and if it is, do we need to prove the statement, or is it rather obvious?
%The lemma implies another algorithm to compute distances (it would not give shortest paths explicitly, unlike alg:find_path)
\begin{lemma}
    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, which means that the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $i$ in tree $T_i$.
\end{lemma}

\begin{proof}
    Induction on $k = d(T_1,T_2)$:

    Basis($k = 1$):
    If rank of $p_{n+1}$ changes between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = 0$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 0 \leq 1 - 1 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
    If on the other hand the rank of $p_{n+1}$ does not change between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = d(T_1,T_2) = 1$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 1 \leq 1 - 0 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.

    Step($k \to k+1$):
    Let $T_1$ ad $T_2$ be trees on $n+1$ taxa with distance $d(T_1,T_2) = k$.
    Let $T_2'$ be the last tree on a shortest path from $T_1$ to $T_2$, i.e. $d(T_1,T_2') = k$.
    We can apply the induction hypothesis on $T_1$ and $T_2'$ and see that $d(T_1|_n, T_2'|_n) \leq d(T_1, T_2') - |r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| := k - m$.
    Now we differ two cases:
    \begin{enumerate}
        \item $p_{n+1}$ changes its rank between $T_2$ and $T_2'$.

        Then the trees $T_2|_n$ and $T_2'|_n$ are equal. \todo{Is this obivous?}
        Therefore, it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m$.
        Since the rank of $p_{n+1}$ changes by one between $T_2$ and $T_2'$ it is either $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m+1$ or $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m-1$.
        In either case, we have $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m = (k+1) - (m-1) \leq d(T_1, T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.

        \item $p_{n+1}$ has same rank in $T_2$ and $T_2'$.

        In this case it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) + 1 \leq (k-m)+ 1$.
        It also follows that $|r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| = |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ and therefore, we can conclude $d(T_1|_n,T_2|_n) \leq k-m+1 = k+1-m = d(T_1,T_2) -  |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
    \end{enumerate}
\end{proof}
